\bf Adaptive Mesh Refinement for 1-D Hyperbolic PDEs

### 2015-03-09 paper_draft1.tex
paper -correct some errors -add some comments

\section{The AMR algorithm}
\noindent
In this paper, we consider only one-dimensional systems. The coarse grid consists of a collection of intervals or cells that cover the entire domain, while refinements cover a subset of the domain and have finer cell-widths. If the cell-width in a fine grid is refined by an even integer $R_L$, then the time-step is also refined by the same factor, so that the ratio $\Delta t/\Delta x$ remains unchanged and for every single time-step on the coarse grid, $R_L$ time steps must be taken on the fine grid.

An error estimation procedure based on Richardson extrapolation, described in more detail in section \ref{whentorefine}, is used to flag the cells which need refinement. Flagged cells are clustered into patches, and a fine grid is generated for each patch (including a buffer zone). A finite volume method in wave-propagation form is used to advance the solution on the coarse grid over a single time-step, and then the solution on the fine grid(s) is advanced over $R_L$ time steps to catch-up.

Every $K$ time steps, the fine grids are regenerated, and the error estimation procedure is re-applied to capture the movement of the discontinuities. The refinement and solution process is repeated.

Each of the above steps is described in more detail below.

@@

Consider a linear system
\[
q_t + A(x) q_x = 0,
\]
where $q \in \reals^m$ and $A \in \reals^{m \times m}$ is diagonalizable with real eigenvalues. On the cell $[x_{i-1/2},x_{i+1/2}$, the matrix $A(x)$ is approximated by a matrix $A_{i-1}$ with eigenvalues $\lambda^1_{i-1}, \ldots, \lambda^m_{i-1}$. Then for the Riemann problem at $x_{i-1/2}$, we decompose $q_i - q_{i-1}$ into waves $\mc{W}^p_{i}, p = 1,2,\ldots,m$ in such a way that the left-going waves are eigenvectors of the matrix $A_{i-1}$, traveling with speeds $\lambda^p_{i-1} <0$, while the right-going waves are eigenvectors of $A_i$ traveling with speeds $\lambda^p_i>0$. The update formula is then
\beq \label{waveprop}
q_i^{n+1} = q_i^n - \frac{k}{h}
\sum\limits_{p=1}^m \left[ (\lambda_i^p)^+ \mc{W}_i^p + (\lambda_{i+1}^p)^- \mc{W}_{i+1}^p\right]
- \frac{k}{h} \left( \tilde{F}_{i+1} - \tilde{F}_i \right)
\eeq
where $k$ is the length of the time-step and $h$ is the cell-width in the spatial grid, and for any real number $a$, $a^+ = \max\{a,0\}$ and $a^- = \max\{-a,0\}$.

The summation term gives the first-order upwind method, while the fluxes $\tilde{F}_i$ are the second-order corrections defined by
\beq
\tilde{F}_i = \frac{1}{2} \sum\limits_{p=1}^m
\left| \lambda_i^p\right| \left( 1-\frac{k}{h} \left| \lambda_i^p\right| \right) \tilde{\mc{W}}_{i+1}
\eeq
where $\tilde{W}_{i+1}^p$ is some limited version of the wave $\mc{W}_i^p$.
When $\tilde{W_i} = W_i$, the method is reduced to upwind + second order correction, i.e. Lax-Wendroff method.

This method is extended to nonlinear systems using the Roe approximate Riemann solver, which linearizes the problem at each cell-interface in such a way that conservation is guaranteed. This method is symbolically written as
\begin{equation*}
q_i^{n+1} = q_i^n - \frac{k}{h}
\sum\limits_{p=1}^m \left[ \mc{A}^+\Delta(q_{i}) + \mc{A}^-\Delta(q_{i+1})\right]
- \frac{k}{h} \left( \tilde{F}_{i+1} - \tilde{F}_i \right)
\end{equation*}
where $\mc{A}^+\Delta(q_{i})$ represents the right-going fluctuation from the left interface of cell $i$, while $\mc{A}^-\Delta(q_{i})$ is the left-going fluctuation from the right interface of cell $i$.

\subsection{determining refinement - error estimation}\label{whentorefine}
The idea behind the refinement step comes from first determining where a refinement may be useful.  
If the solution $q(x,t)$ is smooth enough, we assume that method $D$ (application of eqn(\ref{hires})) has accuracy $O(s)$ in both 
space and time. The local truncation error is $q(x,t+k) -D q(x,t) = \tau(x,t) + k O(k^{s+1}+h^{s+1}))$ where $\tau$ is the leading error.
Taking two time steps suggests $q(x,t+2k) -D^2 q(x,t) = 2\tau(x,t) + k O(k^{s+1}+h^{s+1}))$.
Now, coarsen the grid to have space and time widths $2h$, $2k$ and let the differencing scheme be called $D_{2h}$ in this case. The local truncation error in this case reads $q(x,t+2k) -D_{2h} q(x,t) = 2^{s+1}\tau(x,t) + O(h^{s+2})$. Thus, an error growth estimate at time $t$ can be formed by taking the difference in the error taking two steps with the regular integration scheme, and a single large step using every other grid point:
\begin{equation}
\frac{D^2 q(x,t) - D_{2h} q(x,t)}{2^{s+1}-2} = \tau(x,t) + O(h^{s+2}) \, .
\end{equation}
The difference between the solutions obtained on the two grids at each point is proportional to the local truncation error at that point. At coarse cells where the difference between the two sets of values exceed some tolerance, all four cells contained in the real grid are flagged as requiring refinement. See \cite{berg-coll} for more details. {\it Clustering} is simply to fill gaps in neighboring cells flagged for refinement. Should there be a cell not flagged for refinement between cells that require refinement, the cell in between will be refined as well because in the next step it is likely refinement will be required. A buffer zone around the flagged cells is added to track features of interest in the refined region. This buffer zone must be affiliated with the number of time steps evaluated at a particular level. If a patch at level $L$ is refined by even integer $R_L$, then the time-step is equally refined implying that $R_L$ steps must be taken on the refined grid at level $L+1$ for each step on grid at level $L$.

\subsection{assigning values to new fine grid points}
Once regions where a fine grid are determined, values must be assigned to the fine grid points. Space-time interpolation is needed to supply values for the fine grid because more time steps are taken on the fine grid  than the coarse grid and there are no coarse grid values available at intermediate times. 
This is achieved through a linear interpolation step with known coarse grid values and these values are available since we advance the coarse grid first as stated in the algorithm.

\subsection{flux updates on refined and coarse meshes}
After finishing the recursive step creating the fine grid structure, how to evolve the newly refined grid needs to be considered. To evolve, conservation of fluxes at grid interfaces is required. We refer to Fig \ref{fig21} for this discussion. There are essentially three cases to consider: coarse cell to coarse cell, fine cell to fine cell, and overlapping coarse cell and fine cell regions. Let $F_i$ be the flux at the left edge of coarse cell $i$. This is different notation than what we typically wrote in class, $F_{i-\frac{1}{2}}$. 
\begin{figure}[h]
    \centering
    \includegraphics[width=.6\textwidth]{berg-lev-fig21}
   \caption{Essential figure for flux adapting Cartesian grids. The top grid is the normal coarse grid we are used to using with fluxes indicated. The middle diagram shows the case where a fine grid overlays a coarse grid and the interface region of fine and coarse grids. Recall that $2\hat{h}=h$ and $2\hat{k}=k$. The bottom diagram depicts the ghost cell region imposed on the coarse cell interface as well as the fine grid fluxes needed to bridge the interface.}
    \label{fig21}
\end{figure}

Updating between coarse grid cells is, to lowest order, simply:
\begin{equation}
q_{i}^{1} = q_{i}^{0} - \frac{k}{h} (F_{i+1}^{0} - F_{i}^{0}) 
\end{equation}
\noindent where $h,k$ are the space and time lattice spacing respectively.

In regions that contain $m$ adjacent single level refined cells let the lattice spacing be denoted by $\hat{h},\hat{k}$.
We have, in the stated order, $\forall i=1,m$
\begin{equation}
\begin{array}{c}
\hat{q}_{i}^{1} = \hat{q}_{i}^{0} - \frac{\hat{k}}{\hat{h}} (\hat{F}_{i+1}^{0} - \hat{F}_{i}^{0}) \\
\hat{q}_{i}^{2} = \hat{q}_{i}^{1} - \frac{\hat{k}}{\hat{h}} (\hat{F}_{i+1}^{1} - \hat{F}_{i}^{1}) 
\end{array}
\end{equation}
\noindent in the refined cells where we assume each refinement is $2X$.

In regions where the coarse and fine grids overlap, care must be taken. Two ghost cells are introduced, $\hat{q}_{m+1},\hat{q}_{m+2}$. The values in these 
cells are decided by evaluating a polynomial (determined by interpolating the coarse values $q_{i}^{0}, q_{i}^{1}$)  at these fine grid locations. At the final time (completed time step)  $q_{i}^{1}$ is replaced by the average of the 
fine grid values in the overlapped region:
\begin{equation}
q_{j-1}^{1} = \frac{1}{2} (\hat{q}_{m+1}^{2}+\hat{q}_{m+2}^{2})
\end{equation}
\noindent and the corresponding left edge flux is similarly modified so that $q_{i}^{1}$ is finally modified as:
\begin{equation}\label{fluxcorrect}
q_{i}^{1} \rightarrow q_{i}^{1} + \frac{k}{h} ( \frac{1}{2}(\hat{F}_{m+1}^{0} + \hat{F}_{m+1}^{1}) - F_{i}^{0}) \, .
\end{equation}

\subsection{updating grid interfaces with wave propagation -correction step}
With flux-differencing to ensure conservation, a new coarse grid value is updated by an average of fine-grid values in any cell covered by a fine grid. The key to conservation is that flux into coarse cells is accounted for by flux out of adjacent fine cells, and vice versa, as described in the previous section. Wave propagation algorithms are written in a manner that they can be extended to hyperbolic problems not in conservation form retaining wave features but with no flux function. 
For wave-propagation algorithms, a similar correction is needed for the waves to yield conservation.  The wave-propagation form can be used to update the values at the interface between fine and coarse grid cells on each grid independently using Òghost-cellÓ values as needed near grid interfaces.

Recall from class that $A^+ \Delta q_{i-\frac{1}{2}}$ measures the net effect of all right going waves present from $x_{i-\frac{1}{2}}$, and $A^- \Delta q_{i-\frac{1}{2}}$ measures the net effect of all left going waves from the same interface. We call these {\it fluctuations}.
Whether conservative or non-conservative systems are treated, second-order corrections are still written as flux-differences. 
However, the first-order upwind terms written in terms of the fluctuations must be handled differently. 
To maintain conservation we must {\it first} solve the Riemann problem between the ghost cell value $\hat{q}_{m+1}^{0}$ on the fine grid and the coarse grid value $q_{j}^{0}$, and add to the coarse cell value $q_{j}^{1}$ the resulting total fluctuation (dropped indices) $A^{-}\Delta q + A^{+}\Delta q$ weighted by $\hat{k}/h = 1/2$ since the time step is $\hat{k}$ while the cell size is $h$. 
{\it Second} we must also solve a Riemann problem on the fine grid between $\hat{q}_{m+1}^{1}$ and $q_{j}^{0}$
and add these fluctuations into $q_{j}^{1}$.

The correction step extends to an arbitrary hyperbolic system (i.e. see \ref{linsys} for linear hyperbolic system of PDEs) where we use a Riemann solver to obtain $A^{-}\Delta q$ and  $A^{+}\Delta q$ from the two states $\hat{q}_{m+1}^{0}$ and $q_{j}^{0}$.  This step is made in each of the $R$ time steps on the refined grid within the single coarse-grid step, where $R$ is the refinement ratio. \\

>>>>

\section{The AMR algorithm}
\noindent In this paper, we consider only one-dimensional systems. The coarse grid consists of a collection of intervals or cells that cover the entire domain, while refinements cover a subset of the domain and have finer cell-widths. If the cell-width in a fine grid is refined by an even integer $R_L$, then the time-step is also refined by the same factor, so that the ratio $\Delta t/\Delta x$ remains unchanged and for every single time-step on the coarse grid, $R_L$ time steps must be taken on the fine grid.  
An error estimation procedure based on Richardson extrapolation, described in more detail in section \ref{whentorefine}, is used to flag the cells which need refinement. Flagged cells are clustered into patches, and a fine grid is generated for each patch (including a buffer zone). A finite volume method in wave-propagation form is used to advance the solution on the coarse grid over a single time-step, and then the solution on the fine grid(s) is advanced over $R_L$ time steps to catch-up. 
Every $R$ time steps, the fine grids are regenerated, and the error estimation procedure is re-applied to capture the movement of the discontinuities. The refinement and solution process is repeated. Each of the above steps is described in more detail below.

@@

\noindent Consider a linear system $q_t + A(x) q_x = 0$,
where $q \in \reals^m$ and $A \in \reals^{m \times m}$ is diagonalizable with real eigenvalues. On the cell $[x_{i-1/2},x_{i+1/2}]$, the matrix $A(x)$ is approximated by a matrix $A_{i-1}$ for cell $i-1$ with eigenvalues $\lambda^1_{i-1}, \ldots, \lambda^m_{i-1}$. Then for the Riemann problem at $x_{i-1/2}$, we decompose $q_i - q_{i-1}$ into waves $\mc{W}^p_{i}\, , p = 1,\ldots,m$ in such a way that the left-going waves are eigenvectors of the matrix $A_{i-1}$, traveling with speeds $\lambda^p_{i-1} <0$, while the right-going waves are eigenvectors of $A_i$ traveling with speeds $\lambda^p_i>0$. The update formula is then
\beq \label{waveprop}
q_i^{n+1} = q_i^n - \frac{k}{h}
\sum\limits_{p=1}^m \left[ (\lambda_i^p)^+ \mc{W}_i^p + (\lambda_{i+1}^p)^- \mc{W}_{i+1}^p\right]
- \frac{k}{h} \left( \tilde{F}_{i+1} - \tilde{F}_i \right)
\eeq
where $k$ is the length of the time-step and $h$ is the cell-width in the spatial grid, and for any real number $a$, $a^+ = \max\{a,0\}$ and $a^- = \max\{-a,0\}$.

\noindent The summation term gives the first-order upwind method, while the fluxes $\tilde{F}_i$ are the second-order corrections defined by
\beq
\tilde{F}_i = \frac{1}{2} \sum\limits_{p=1}^m
\left| \lambda_i^p\right| \left( 1-\frac{k}{h} \left| \lambda_i^p\right| \right) \tilde{\mc{W}}_{i+1}
\eeq
where $\tilde{W}_{i+1}^p$ is some limited version of the wave $\mc{W}_i^p$.
When $\tilde{W_i} = W_i$, the method is reduced to upwind + second order correction, i.e. Lax-Wendroff method.

\noindent This method is extended to nonlinear systems using the Roe approximate Riemann solver, which linearizes the problem at each cell-interface in such a way that conservation is guaranteed. This method is symbolically written as
\begin{equation*}
q_i^{n+1} = q_i^n - \frac{k}{h}
\sum\limits_{p=1}^m \left[ \mc{A}^+\Delta(q_{i}) + \mc{A}^-\Delta(q_{i+1})\right]
- \frac{k}{h} \left( \tilde{F}_{i+1} - \tilde{F}_i \right)
\end{equation*}
where $\mc{A}^+\Delta(q_{i})$ represents the right-going fluctuation from the left interface of cell $i$, while $\mc{A}^-\Delta(q_{i})$ is the left-going fluctuation from the right interface of cell $i$.

\subsection{determining refinement - error estimation}\label{whentorefine}
The idea behind the refinement step comes from first determining where a refinement may be useful.  
If the solution $q(x,t)$ is smooth enough, we assume that method $D$ (application of eqn(\ref{waveprop})) has accuracy $O(s)$ in both 
space and time. The local truncation error is $q(x,t+k) -D q(x,t) = \tau(x,t) + k O(k^{s+1}+h^{s+1}))$ where $\tau$ is the leading error.
Taking two time steps suggests $q(x,t+2k) -D^2 q(x,t) = 2\tau(x,t) + k O(k^{s+1}+h^{s+1}))$.
Now, coarsen the grid to have space and time widths $2h$, $2k$ and let the differencing scheme be called $D_{2h}$ in this case. The local truncation error in this case reads $q(x,t+2k) -D_{2h} q(x,t) = 2^{s+1}\tau(x,t) + O(h^{s+2})$. Thus, an error growth estimate at time $t$ can be formed by taking the difference in the error taking two steps with the regular integration scheme, and a single large step using every other grid point:
\begin{equation}
\frac{D^2 q(x,t) - D_{2h} q(x,t)}{2^{s+1}-2} = \tau(x,t) + O(h^{s+2}) \, .
\end{equation}
The difference between the solutions obtained on the two grids at each point is proportional to the local truncation error at that point. At coarse cells where the difference between the two sets of values exceed some tolerance, all four cells contained in the real grid are flagged as requiring refinement. See \cite{berg-coll} for more details. {\it Clustering} is simply to fill gaps in neighboring cells flagged for refinement. Should there be a cell not flagged for refinement between cells that require refinement, the cell in between will be refined as well because in the next step it is likely refinement will be required. A buffer zone around the flagged cells is added to track features of interest in the refined region. This buffer zone must be affiliated with the number of time steps evaluated at a particular level. If a patch at level $L$ is refined by even integer $R_L$, then the time-step is equally refined implying that $R_L$ steps must be taken on the refined grid at level $L+1$ for each step on grid at level $L$.

\subsection{assigning values to new fine grid points}
Once regions where a fine grid are determined, values must be assigned to the fine grid points. Current time fine grid values are assigned by evaluating a linear interpolation akin to flux limiting between coarse cell values in neighboring cells. Intermediate time values on the fine grid are assigned by evaluating a bilinear space-time interpolation based on coarse grid values between neighboring cells at the current and subsequent time. This approach is needed because more time steps are taken on the fine grid  than the coarse grid and there are no coarse grid values available at intermediate times.

\subsection{flux updates on refined and coarse meshes}
After recursively creating the fine grid, how to evolve the newly refined grid needs to be considered. To evolve, conservation of fluxes at grid interfaces is required. We refer to Fig \ref{fig21} for this discussion. There are essentially three cases to consider: coarse cell to coarse cell, fine cell to fine cell, and overlapping coarse cell and fine cell regions. Let $F_i$ be the flux at the left edge of coarse cell $i$. This is different notation than what we typically wrote in class, $F_{i-\frac{1}{2}}$. 
\begin{figure}[h]
    \centering
    \includegraphics[width=.6\textwidth]{berg-lev-fig21}
   \caption{Essential figure for flux adapting Cartesian grids. The top grid is the normal coarse grid we are used to using with fluxes indicated. The middle diagram shows the case where a fine grid overlays a coarse grid and the interface region of fine and coarse grids. Recall that $2\hat{h}=h$ and $2\hat{k}=k$. The bottom diagram depicts the ghost cell region imposed on the coarse cell interface as well as the fine grid fluxes needed to bridge the interface.}
    \label{fig21}
\end{figure}

\noindent Updating between coarse grid cells is, to lowest order, simply:
\begin{equation}
q_{i}^{1} = q_{i}^{0} - \frac{k}{h} (F_{i+1}^{0} - F_{i}^{0}) 
\end{equation}
\noindent where $h,k$ are the space and time lattice spacing respectively.

\noindent In regions that contain $m$ adjacent single level refined cells let the lattice spacing be denoted by $\hat{h},\hat{k}$.
We have, in the stated order, $\forall i=1,m$
\begin{equation}
\begin{array}{c}
\hat{q}_{i}^{1} = \hat{q}_{i}^{0} - \frac{\hat{k}}{\hat{h}} (\hat{F}_{i+1}^{0} - \hat{F}_{i}^{0}) \\
\hat{q}_{i}^{2} = \hat{q}_{i}^{1} - \frac{\hat{k}}{\hat{h}} (\hat{F}_{i+1}^{1} - \hat{F}_{i}^{1}) 
\end{array}
\end{equation}
\noindent in the refined cells where we assume each refinement is $2X$ as in the figure.

\noindent In regions where the coarse and fine grids overlap, care must be taken. Two ghost cells are introduced, $\hat{q}_{m+1},\hat{q}_{m+2}$. The values in these 
cells are decided by evaluating a polynomial at these fine grid locations determined by interpolating coarse grid values (as stated earlier). To complete the time step,  $q_{i}^{1}$ is replaced by the average of the fine grid values in the overlapped region:
\begin{equation}
q_{j-1}^{1} = \frac{1}{2} (\hat{q}_{m+1}^{2}+\hat{q}_{m+2}^{2})
\end{equation}
\noindent and the corresponding left edge flux is similarly modified so that $q_{i}^{1}$ is finally modified as:
\begin{equation}\label{fluxcorrect}
q_{i}^{1} \rightarrow q_{i}^{1} + \frac{k}{h} ( \frac{1}{2}(\hat{F}_{m+1}^{0} + \hat{F}_{m+1}^{1}) - F_{i}^{0}) \, .
\end{equation}

\subsection{updating grid interfaces with wave propagation -correction step}
With flux-differencing to ensure conservation, a new coarse grid value is updated by an average of fine-grid values in any cell covered by a fine grid. The key to conservation is that flux into coarse cells is accounted for by flux out of adjacent fine cells, and vice versa, as described in the previous section. Wave propagation algorithms are written in a manner that they can be extended to hyperbolic problems not in conservation form retaining wave features but with no flux function. 
For wave-propagation algorithms, a similar correction is needed for the waves to yield conservation.  The wave-propagation formis used to update the values at the interface between fine and coarse grid cells on each grid independently using Òghost-cellÓ values as needed near grid interfaces.

\noindent Recall from class that $A^+ \Delta q_{i-\frac{1}{2}}$ measures the net effect of all right going waves present from $x_{i-\frac{1}{2}}$, and $A^- \Delta q_{i-\frac{1}{2}}$ measures the net effect of all left going waves from the same interface. We call these {\it fluctuations}.
Whether conservative or non-conservative systems are treated, second-order corrections are still written as flux-differences. 
However, the first-order upwind terms written in terms of the fluctuations must be handled differently. 
To maintain conservation we must {\it first} solve the Riemann problem between the ghost cell value $\hat{q}_{m+1}^{0}$ on the fine grid and the coarse grid value $q_{j}^{0}$, and add to the coarse cell value $q_{j}^{1}$ the resulting total fluctuation (dropped indices) $A^{-}\Delta q + A^{+}\Delta q$ weighted by $\hat{k}/h = 1/2$ since the time step is $\hat{k}$ while the cell size is $h$. 
{\it Second} we must also solve a Riemann problem on the fine grid between $\hat{q}_{m+1}^{1}$ and $q_{j}^{0}$
and add these fluctuations into $q_{j}^{1}$.

\noindent The correction step extends to an arbitrary hyperbolic system (i.e. see \ref{linsys} for linear hyperbolic system of PDEs) where we use a Riemann solver to obtain $A^{-}\Delta q$ and  $A^{+}\Delta q$ from the two states $\hat{q}_{m+1}^{0}$ and $q_{j}^{0}$.  This step is made in each of the $R$ time steps on the refined grid within the single coarse-grid step, where $R$ is the refinement ratio. \\
